### 2016-04-21 src/3_TimeSeries.tex
chapter 2 reviewed

\subsubsection{Deterministic trend}

Deterministic trend or trend stationary processes have the following form:

\[
X_t = f(t) + \epsilon{t} \quad ,
\]

\noindent where $t$ is the time trend and $\epsilon{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
\begin{itemize}
\item If $f(t) = \alpha + \beta t$ we have a linear trend model which is widely used. 
\item If $f(t)= \alpha \exp^rt$ we have an exponential growth curve.
\item If $f(t) = c_1 + c_2t + c_3 t^2$ we have a quadratic trend model
\item If $f(t) = \frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
\end{itemize}

\subsubsection{Stochastic trend}
Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:

\[
X_t = \mu + X_{t-1} + \epsilon_t
\]
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.

@@

\[
(1-L) X_t = \mu  + \epsilon_t
\]

@@

Other less common forms of non-stationarity are structural break in mean and structural break in variance.

\newpage
\subsection{Spurious regression} \label{sec:spurious}
The use
of non-stationary data can lead to spurious regressions. Spurious regression
dates back to Yule in 1926 \cite{yule1926} If two stationary
variables are generated as independent random series and are trending over time,
when one of those variables is regressed on the other, they could have a high
$R^2$ even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under standard
measures but valueless \cite{brooks2002}.

\textbf{Example of spurious regression} \quad 
An example is shown below. Two random walks series $x_1t$ and $x_2t$ with a small drift ($\alpha=0.09$) are generated and
then regressed one on the other:
 \begin{equation}
 x_i,t = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
 \end{equation}

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.8\textwidth]{img/spurious1}
  \caption{Two random walks time series $x_1,t$ and $x_2,t$}
  \label{fig:spurious1}
\end{figure}

@@

%                            OLS Regression Results                            
%==============================================================================
%Dep. Variable:                      y   R-squared:                       0.883
%Model:                            OLS   Adj. R-squared:                  0.882
%Method:                 Least Squares   F-statistic:                     742.2
%Date:                Fri, 06 Feb 2015   Prob (F-statistic):           1.61e-47
%Time:                        06:07:47   Log-Likelihood:                -218.03
%No. Observations:                 100   AIC:                             440.1
%Df Residuals:                      98   BIC:                             445.3
%Df Model:                           1                                         
%==============================================================================
%                 coef    std err          t      P>|t|      [95.0% Conf. Int.]
%------------------------------------------------------------------------------
%x1             0.8883      0.033     27.243      0.000         0.824     0.953
%const          2.6726      0.405      6.606      0.000         1.870     3.475
%==============================================================================
%Omnibus:                        4.133   Durbin-Watson:                   0.363
%Prob(Omnibus):                  0.127   Jarque-Bera (JB):                4.145
%Skew:                          -0.469   Prob(JB):                        0.126
%Kurtosis:                       2.662   Cond. No.                         23.3
%==============================================================================
$\blacksquare$

\newpage
\subsection{Integration}

@@

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

\subsection{Cointegration} \label{sec:coint}
Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.

@@

Stock and Watson in 1988 \cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationships.

@@

There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}).

@@

Where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.  $\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$. $\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration vectors is $r$ against an alternative of $r
+1$.

@@

If we have two-dimensional process $\mathbf{X}_t$, $t=1,\dots,T$ by:

\begin{eqnarray*}
\mathbf{X}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
\mathbf{X}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
\end{eqnarray*}

Since $\mathbf{X}_{1t}$ and $\mathbf{X}_{2t}$ are I(1) processes and there
exist a vector $\beta = [a -1]$ such that:

\[
\beta^\top \mathbf{X}_t = a \mathbf{X}_{1t} -\mathbf{X}_{2t} = 
a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
\]

@@

If we add a I(0) process
$\mathbf{X}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
&0&1\end{bmatrix}$ since:

\[
\beta^\top \mathbf{X}_t = 
\begin{bmatrix}
a & -1 & 0 \\
0 & 0 & 1
\end{bmatrix} 
\begin{bmatrix} 
\mathbf{X}_{1t} \\
\mathbf{X}_{2t} \\
\mathbf{X}_{3t}
\end{bmatrix} = 
\begin{bmatrix}
a\epsilon_{2t} - \epsilon_{3t} \\
\epsilon_{4t}
\end{bmatrix}
\]

\begin{figure}[!h]
  \centering
  \includegraphics[width=\textwidth]{img/coint-example}
  \caption{Cointegration example: two processes I(1)(blue and green) and one process I(0) (red)}
  \label{fig:cointex}
\end{figure}

@@

$\blacksquare$

\section{Unit root tests}
Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance. An important econometric task is to determine the type of non-stationary. For trend stationary I(0) time series a time-trend regression is appropriate. For I(1) time series taking first differences transforms it into a stationary one.  For cointegration to determine if time series are I(1) is mandatory to model their long-run relationship. In order to determine if a time series is an a I(1) or I(0) process unit root tests are used. Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed in 1984 by Said and Dickey in 1984\cite{said1984}.

@@

The volatility of the data is due to a large number of factors affecting the
market, with some of them directly measurable such as historical prices,
trends, supply and demand. However, others such as monetary policies and news
are not directly measurable and are included in other models out of the scope
of this thesis.

@@

The volatility of a stock is not directly observable~\cite{tsay2005,engle1993}. For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.  If intraday data is available, then volatility could be estimated. However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed. These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}. Examples of volatility proxies are the following:

\subsubsection{realized volatility} is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realized volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realized
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\subsubsection{implied volatility volatility} not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.

@@

The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy to use, successful and flexible models for the analysis of multivariate time series. It is a natural extension of the univariate autoregressive (AR) model. The VAR model has proven to be useful for describing the dynamic behaviour of economic and financial time series.
VAR is a general framework describing the behaviour of a
set of $l$ endogenous variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{\lambda,t}$, $\lambda=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

@@

Notice that, regarding our notation of section (\ref{sec:coint}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
$\lambda=1,\dots,l$.
Transposing each equation of the system (\ref{eq:var}) we can write
the VAR($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
\end{equation}
%%
\noindent where:
%%
\begin{alignat}{2}
\mathbf{Y}
&= \begin{bmatrix}
   \mathbf{y}_{p+1}^\top \\
   \mathbf{y}_{p+2}^\top \\
   \vdots \\
   \mathbf{y}_N^\top
   \end{bmatrix}_{(N-p)\times l}
&\quad
\mathbf{A}
&= \begin{pmat}[{...|}]
   \mathbf{y}_p^\top & \mathbf{y}_{p-1}^\top & \dots 
                    & \mathbf{y}_1^\top & 1 \cr
   \mathbf{y}_{p+1}^\top & \mathbf{y}_p^\top & \dots
                       & \mathbf{y}_2^\top & 1 \cr
   \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top & \mathbf{y}_{N-2}^\top & \dots 
                       & \mathbf{y}_{N-p}^\top & 1 \cr
   \end{pmat}_{(N-p)\times(pl+1)} \\
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\Phi}_1^\top \\
   \boldsymbol{\Phi}_2^\top \\
   \vdots \\
   \boldsymbol{\Phi}_p^\top \\
   \mathbf{c}^\top
   \end{bmatrix}_{(pl+1)\times l}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}_{(N-p)\times l}
\end{alignat}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation.

\section{Vector error correction model}
Vector error correction model (VECM)  \cite{engle87} is a special form of a VAR model for I(1) variables that are also
cointegrated~\cite{banerjee1993}.

>>>>

\begin{enumerate}[(a)]
\item \textbf{Deterministic trend:} Deterministic trend or trend stationary processes have the following form:

\[
y_t = f(t) + \epsilon_{t} \quad ,
\]

\noindent where $t$ is the time trend and $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time:
\begin{itemize}
\item If $f(t) = \alpha + \beta t$ we have a linear trend model which is widely used. 
\item If $f(t)= \alpha \exp^rt$ we have an exponential growth curve.
\item If $f(t) = c_1 + c_2t + c_3 t^2$ we have a quadratic trend model
\item If $f(t) = \frac{1}{k+\alpha \beta ^t }$ we have a logistic curve
\end{itemize}
\item \textbf{Stochastic trend:} Stochastic trend processes are also called unit root or difference stationarity processes and have the following form:

\[
y_t = \mu + y_{t-1} + \epsilon_t
\]
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.

@@

\[
(1-L) y_t = \mu  + \epsilon_t
\]

@@

Other less common forms of non-stationarity are structural break in mean and structural break in variance. 
\end{enumerate}

\subsection{Spurious regression} \label{sec:spurious}
The use of non-stationary data can lead to spurious regressions. Spurious regression
dates back to Yule in 1926 \cite{yule1926}. If two stationary
variables are generated as independent random series and are trending over time,
when one of those variables is regressed on the other, they could have a high
$R^2$ even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under standard
measures but valueless \cite{brooks2002}.

\textbf{Example of spurious regression} \quad 
An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with a small drift ($\alpha=0.09$) are generated and
then regressed one on the other:
 \begin{equation}
 x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad \mathcal{N}(0,1), \quad i=1,2
 \end{equation}

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.8\textwidth]{img/spurious1}
  \caption{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
  \label{fig:spurious1}
\end{figure}

@@

\subsection{Integration}

@@

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation a stationary process is
obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

\subsection{Cointegration} \label{sec:cointegration}
Cointegration concept was introduced by Engle in 1987 \cite{engle1987} and implies that one or
more linear combinations of non-stationary variables are stationary even though
individually they are not.

@@

Stock and Watson in 1988 \cite{stock+watson1988} observed that
cointegration reflects the common stochastic trends providing a useful way to
understand cointegration relationships. These common stochastic trends can be
also interpreted as a long-run equilibrium relationship.

@@

There are two statistics for cointegration under Johansen approach: the trace (shown in equation \ref{eq:johansentrace}) and the maximum-eigenvalue statistic (shown in equation \ref{eq:johansenmax}):

@@

\noindent where $r$ is the number of cointegration vectors and $\hat]{\lambda}_i$ is the estimated value for the ith ordered eigenvalue.  $\lambda_{\text{trace}}$ null hypothesis is that the number of cointegration vectors is less than or equal to $r$ against that there are more than $r$. $\lambda_{\text{max}}$ has the null hypothesis that the number of cointegration vectors is $r$ against an alternative of $r
+1$.

@@

If we have two-dimensional process $\mathbf{y}_t$, $t=1,\dots,T$ by:

\begin{eqnarray*}
\mathbf{y}_{1t} &=& \sum_{i=1}^t \epsilon_{1i} + \epsilon_{2t} \\
\mathbf{y}_{2t} &=& a \sum_{i=1}^t \epsilon_{1i} + \epsilon_{3t} 
\end{eqnarray*}

Since $\mathbf{y}_{1t}$ and $\mathbf{y}_{2t}$ are I(1) processes and there
exist a vector $\beta = [a -1]$ such that:

\[
\beta^\top \mathbf{y}_t = a \mathbf{y}_{1t} -\mathbf{y}_{2t} = 
a\epsilon_{2t} - \epsilon_{3t} \sim \text{I(0)}
\]

@@

If we add a I(0) process
$\mathbf{y}_{3t} = \epsilon_{4t}$  we find that there exists two cointegration
vectors now: $\begin{bmatrix}a &-1& 0\end{bmatrix}$ and $\begin{bmatrix}0
&0&1\end{bmatrix}$ since:

\[
\beta^\top \mathbf{y}_t = 
\begin{bmatrix}
a & -1 & 0 \\
0 & 0 & 1
\end{bmatrix} 
\begin{bmatrix} 
\mathbf{y}_{1t} \\
\mathbf{y}_{2t} \\
\mathbf{y}_{3t}
\end{bmatrix} = 
\begin{bmatrix}
a\epsilon_{2t} - \epsilon_{3t} \\
\epsilon_{4t}
\end{bmatrix}
\]

\begin{figure}[!h]
  \centering
  \includegraphics[width=\textwidth]{img/coint-example}
  \caption{Cointegration example: two I(1) processes (blue and green) and one I(0) process (red).}
  \label{fig:cointex}
\end{figure}

@@

\section{Unit root tests}
Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance. An important econometric task is to determine the type of non-stationary. For trend stationary I(0) time series a time-trend regression is appropriate. For I(1) time series taking first differences transforms it into a stationary one.  For cointegration to determine if time series are I(1) is mandatory to model their long-run relationship. In order to determine if a time series is an a I(1) or I(0) process unit root tests are used. Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984}.

@@

The volatility of the data is due to a large number of factors affecting the
market, with some of them directly measurable such as historical prices,
trends, supply and demand. However, others such as monetary policies and news
are not directly measurable.

@@

The volatility of a stock is not directly observable \cite{tsay2005,engle1993}. For example, daily volatility is not directly observable from only daily returns because there is only one observation in a trading day.  If intraday data is available, then volatility could be estimated. However, intraday returns are not the only explanatory variables for volatility and several estimators have been proposed. These estimators are observable variables that are related to the latent variable of interest called volatility proxies~\cite{devilderetal2007}. Examples of volatility proxies are the following:

\subsubsection{Realized volatility} is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realized volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realized
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\subsubsection{Implied volatility} not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.

@@

The vector autoregressive VAR($p$) model \cite{sims1980} is one of the most easy to use, successful and flexible models for the analysis of multivariate time series. It is a natural extension of the univariate autoregressive (AR) model. The VAR model has proven to be useful for describing the dynamic behaviour of economic and financial time series.
VAR is a general framework describing the behaviour of a
set of $l$ endogenous variables as a linear combination of their last $p$
values, where $l,p\in\mathbb{N}$. 
In our case, each one of these $l$ variables is a scalar time series
$y_{i,t}$, $i=1,\dots,l$, and we represent them all together
at time $t$ by the the vector time series:
\begin{equation*}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation*}
\noindent
Notice that the vector $\mathbf{y}_t$ is assumed to be $l$-dimensional.

@@

Notice that, regarding our notation of section (\ref{sec:cointegration}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $i$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{i,t}^\nu$, where $\nu=1,\dots,p$ and
$i=1,\dots,l$.
Transposing each equation of the system (\ref{eq:var}) we can write
the VAR($p$) model in block-matrix form as:
\begin{equation}\label{eq:simplevareq}
\mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
\end{equation}
%%
\noindent where:
%%
\begin{alignat*}{2}
\mathbf{Y}
&= \begin{bmatrix}
   \mathbf{y}_{p+1}^\top \\
   \mathbf{y}_{p+2}^\top \\
   \vdots \\
   \mathbf{y}_N^\top
   \end{bmatrix}_{(N-p)\times l}
&\quad
\mathbf{A}
&= \begin{pmat}[{...|}]
   \mathbf{y}_p^\top & \mathbf{y}_{p-1}^\top & \dots 
                    & \mathbf{y}_1^\top & 1 \cr
   \mathbf{y}_{p+1}^\top & \mathbf{y}_p^\top & \dots
                       & \mathbf{y}_2^\top & 1 \cr
   \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top & \mathbf{y}_{N-2}^\top & \dots 
                       & \mathbf{y}_{N-p}^\top & 1 \cr
   \end{pmat}_{(N-p)\times(pl+1)} \\
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\Phi}_1^\top \\
   \boldsymbol{\Phi}_2^\top \\
   \vdots \\
   \boldsymbol{\Phi}_p^\top \\
   \mathbf{c}^\top
   \end{bmatrix}_{(pl+1)\times l}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}_{(N-p)\times l}
\end{alignat*}
Taking into account the error term $\mathbf{E}$, equation \ref{eq:simplevareq}
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation (see section \ref{sec:OLS}).

\section{Vector error correction model}
Vector error correction model (VECM)  \cite{engle87} describe the joint behaviour of a set of cointegrated I(1) time series \cite{banerjee1993} and can be derived from the simple VAR model.

It is obtained re-writing equation \ref{eq:var} in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of the model VAR shown in equation \ref{eq:var}, are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
nonsingular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:simpleveceq}
\mathbf{Y} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
